---
title: "Simulated data used for training ML models in the PlausibleAlienZoo study"
author: "Ulrike Kuhl"
date: "7/10/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This code simulates datasets with known dependencies between features for the AlienZoo study line.
Written by Ulrike Kuhl; for comments, questions etc.: kuhl@techfak.uni-bielefeld.de.

# Let's start simulating

```{r results='asis'}
# import required modules
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
library(ggplot2)
library(dplyr)
library(data.table)

```
Define basic properties of features and variables.

```{r}
# minimal / maximal growth rates:
GRneg.min=0.1
GRneg.max=1.0
GRpos.min=1.1
GRpos.max=1.9

F.min=0
F.max=7
```

## Data set fir the Plausible Alien Zoo study: simple linear relationship between one feature and the outcome, depending on the value of either one of two additional variables

* 3276800 data points (given by combinatoric nature of having 5 features that may take one of 8 values = 3276800 combinations; each combination being repeated 100 times); 
* there is a linear relationship between plant 2 and the growth rate, iff
  + plant 4 has a value of 1 or 2 OR
  + plant 5 is not smaller than 4

```{r}
# define the possible values each feature might take
featureVals.possible=c(F.min:F.max)
# expand to have a set with 5 features and all possible combinations of features
df.set=expand.grid(featureVals.possible,featureVals.possible,featureVals.possible,
                   featureVals.possible,featureVals.possible)
# repeat to have each combination 100 times
df.set=df.set[rep(seq_len(nrow(df.set)), each = 100), ]

# generate new field for growth rate values
# growth rate values per default in a negative range
set.seed(42)
df.set$GR=round(runif(nrow(df.set), min=GRneg.min, max=GRneg.max),2)

# transform GR values of deciding feature 'plant 2', 
# depending on the values of 'plant 4' and 'plant 5': 
df.set$GR[df.set$Var4==1 | df.set$Var4==2 | df.set$Var5>3 ]=
  df.set$Var2[df.set$Var4==1 | df.set$Var4==2 | df.set$Var5>3 ]

OldRange = (F.max - F.min)  
NewRange = (GRpos.max - GRpos.min) 
df.set$GR[df.set$Var4==1 | df.set$Var4==2 | df.set$Var5>3] = 
  jitter((((df.set$GR[df.set$Var4==1 | df.set$Var4==2 | df.set$Var5>3] - F.min) * NewRange) / OldRange) 
         + GRpos.min, factor=1)

# destroy relationship for variable values 6 and 7 (= 'overfeeding')
OldRange = (F.max - F.min)
NewRange = (GRneg.max - GRneg.min)
# note: depending on var1 here, to destroy any dependency with var 2
df.set$GR[df.set$Var2==6]=round(jitter((((df.set$Var1[df.set$Var2==6] - F.min) * NewRange) / OldRange) 
                                       + GRneg.min, factor = 1),2)
df.set$GR[df.set$Var2==7]=round(jitter((((df.set$Var1[df.set$Var2==7] - F.min) * NewRange) / OldRange) 
                                       + GRneg.min, factor = 1),2)

# take subset of data frame for plotting so it does not take too long:
df.set.plot=df.set
set.seed(42)
rows=sample(nrow(df.set.plot))
df.set.plot <- df.set.plot[rows, ]
df.set.plot = df.set.plot[seq(1, nrow(df.set), 1000), ]
names(df.set.plot)=c("Plant 1","Plant 2","Plant 3","Plant 4","Plant 5","GrowthRate")

p = df.set.plot %>% tidyr::gather("id", "value", 1:5) %>% 
  ggplot(., aes(GrowthRate, value))+
  geom_point(shape=4, alpha=0.5, size=0.5)+
  facet_wrap(~id,scales="free")

p

# save data!
##write.csv(df.set,'AlienZooDataSet_PAZ.csv', row.names = FALSE)
```

Visualize dependencies as a pretty plot, suitable for a potential publication:

```{r}

library(GGally)
library(ggpubr)

set.seed(42)
rows=sample(nrow(df.set))
tmp <- df.set[rows, ]
tmp = tmp[seq(1, nrow(df.set), 500), ]
names(tmp)=c("Plant 1","Plant 2","Plant 3","Plant 4","Plant 5","GrowthRate")

# function 1: produce jittered scatter plots
my_bin <- function(data, mapping) {
  ggplot(data = data, mapping = mapping) +
    geom_jitter(aes(colour=GrowthRate))+
    scale_color_gradient2(low="#656464", high="#FF2500", midpoint = 1,
                          space = "Lab",guide = "colourbar")+
    theme_bw(base_size = 10)+
    theme(strip.background=element_rect(fill="white"))
}

# function 2: make sure to only keep lower triangle facets in ggpairs
gpairs_lower <- function(g){
  g$plots <- g$plots[-(1:g$nrow)]
  g$yAxisLabels <- g$yAxisLabels[-1]
  g$nrow <- g$nrow -1
  g$plots <- g$plots[-(seq(g$ncol, length(g$plots), by = g$ncol))]
  g$xAxisLabels <- g$xAxisLabels[-g$ncol]
  g$ncol <- g$ncol - 1
  g
}

# re-order data frame; put all GRs > 1 at back so they are plotted last
# (makes pattern in data more easily discernable)
A=data.frame(rbind(tmp[tmp$GrowthRate<1,],tmp[tmp$GrowthRate>=1,]))
names(A)=c("Plant 1","Plant 2","Plant 3","Plant 4","Plant 5","GrowthRate")

# make pairs plot:
pm <- ggpairs(
  data=A, columns = c(1,2,3,4,5),
  lower = list(
    continuous = my_bin
  ),
  upper  = list(continuous = "blank"),
  diag  = list(continuous = "blankDiag"),
  #legend = 1,
  switch="both"
)

pm=gpairs_lower(pm)

# make auxplot to get legend:
auxplot=ggplot(data = A) +
    geom_jitter(aes(x=`Plant 1`,y=`Plant 1`,colour=GrowthRate))+
    scale_color_gradient2(low="#656464", high="#FF2500",midpoint = 1,
                          space = "Lab",guide = "colourbar")
# grab legend:
auxlegend=grab_legend(auxplot)

# add legend:
pm[2,4]=auxlegend

# show
pm

# save
ggsave("DataDist_PAZ.pdf",width = 5, height = 5,)

```

